\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2023-02-27 paper/main.tex
finish model setup

A pairwise comparison process starts with assigning all
individuals of the population the same strategy. Each elementary time step of
the process consists of the mutation phase, the game phase and the update phase.
In the game phase individuals are matched in pairs and that they participate in
a repeated 2 person donation game; a special case of the prisoner's dilemma. In
the donation game there are two actions: cooperation (\(C\)) and defection
(\(D\)). By cooperating a player provides a benefit \(b\) to the other player at
their cost \(c\), with \(0 < c < b\). Thus the payoffs for a player in each turn
are,

@@

Let \(\mathbf{u} = (b-c, -c, b, 0)\) be payoffs in a vector format, and let
\(\mathcal{U} = \{r, s, t, p\}\) denote the set of feasible payoffs, where \(r\)
denotes the payoff of mutual cooperation, \(s\) the sucker's payoff, \(t\) the
temptation to defect payoff, and \(p\) the punishment payoff.

In repeated games there are infinite many strategies, however, similar to the
literature we will assume that individuals use reactive strategies. A reactive
strategy considers only the previous action of the other player, and thus, a
reactive strategy \(s\) can be written as a three-dimensional vector \(s=(y, p,
q)\). The parameter \(y\) is the probability that the strategy opens with a
cooperation and \(p\), \(q\) are the probabilities that the strategy cooperates
given that the opponent cooperated and defected equivalently. The play between a
pair of reactive strategies ($s_1 = (y_1, p_1, q_1)$, $s_2 = (y_2, p_2, q_2)$)
can be model as a Markov process with the transition matrix \(M\),

@@

For the results presented here we assume that mutations are rare (\(\mu
\rightarrow 0\)). In fact, so rare that only two different strategies can be
present in the population at any given time. However, in the Supplementary Information
Section 4 we show in that the main result hold for \(\mu \neq 0\).
The case of low mutation is vastly adopted because it allows us to explicitly
calculate the fixation probability of a newly introduced mutant. More specifically,
at each step
one individual adopts a mutant strategy randomly selected from the set of
feasible strategies. The fixation probability \(\phi_{M}\) of the mutant
strategy can be calculated explicitly,

@@

where \(\lambda^-_k, \lambda^+_k\) are the probabilities that the number of
mutants decreases and increases respectively, and \(k\) is the number of
mutants. Depending on the fixation probability \(\phi_{M}\) the mutant either
fixes (becomes the new resident) or goes extinct. Regardless, in the elementary
time step another mutant strategy is introduced to the population. We iterate
this elementary population updating process for a large number of mutant
strategies and we record the resident strategies at each time step. The
probabilities \(\lambda^-_k \text{ and } \lambda^+_k\) depend on the updating
payoffs of the mutant and the resident strategies. In the next section we
present how they are calculated in the cases of perfect and limited memory.

\subsection{Updating payoffs based Perfect and Limited Memory}

\subsubsection*{Perfect Memory}

In the perfect memory case an individual updates based on the average payoff
against each other member of the population, otherwise as expected payoffs. In
an infinitely repeated game ($\delta \rightarrow 1$) the payoff of a reactive
strategy \(s_1\) against the reactive strategy \(s_2\) can explicitly be
calculated using the stationary vector \(\mathbf{v}\) and the payoffs' vector as
\(\langle\mathbf{v}(s_1,s_2),\mathbf{u}\rangle\). In a population of size \(N\)
there are \(k\) mutants and \(N - k\) residents. We denote the strategies of
a mutant and of a resident as \(s_M =(y_M, p_M, q_M)\) and \(s_R =
(y_R, p_R, q_R)\). The expected payoffs of a resident (\(\pi_R\)) and of a
mutant (\(\pi_M\)) are given by,

@@

The probabilities that the number of mutants decreases and increases,
\(\lambda^-_k\) and \(\lambda^+_k\), in the perfect memory case are defined as,

@@

\subsubsection*{Limited Memory}

In this case of limited memory we initially define the probability that a
reactive strategy receives the payoff $u\!\in\! \mathcal{U}$ in the very last
round of the game. This is given by Proposition~\ref{proposition:last_round}
(see Supplementary Information Section 2.2.1 for proof).

@@

  v_{r}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{y_1y_2}{1\!-\!\delta^2 l_1 l_2}+\delta \frac{\Big(q_1+l_1\big((1\!-\!\delta)y_2+\delta q_2\big)\Big) \Big(q_2+l_2\big((1\!-\!\delta)y_1+\delta q_1\big)\Big)}
  {\displaystyle(1\!-\!\delta l_1l_2)(1\!-\!\delta^2 l_1 l_2)} \times r,\\[1cm]

  v_{s}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{y_1\bar{y}_2}{1\!-\!\delta^2 l_1 l_2}+\delta \frac{\Big(q_1+l_1\big((1\!-\!\delta)y_2+\delta q_2\big)\Big) \Big(\bar{q}_2+\bar{r}_2\big((1\!-\!\delta)y_1+\delta p_1\big)\Big)}
  {\displaystyle(1\!-\!\delta l_1l_2)(1\!-\!\delta^2 l_1 l_2)} \times s,\\[1cm]

  v_{t}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{\bar{y}_1y_2}{1\!-\!\delta^2 l_1 l_2}+\delta \frac{\Big(\bar{q}_1+\bar{r}_1\big((1\!-\!\delta)y_2+\delta p_2\big)\Big) \Big(q_2+l_2\big((1\!-\!\delta)y_1+\delta q_1\big)\Big)}
  {\displaystyle(1\!-\!\delta l_1l_2)(1\!-\!\delta^2 l_1 l_2)} \times t,\\[1cm]

  v_{p}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{\bar{y}_1\bar{y}_2}{1\!-\!\delta^2 l_1 l_2}+\delta \frac{\Big(\bar{q}_1+\bar{r}_1\big((1\!-\!\delta)y_2+\delta p_2\big)\Big) \Big(\bar{q}_2+\bar{r}_2\big((1\!-\!\delta)y_1+\delta p_1\big)\Big)}
  {\displaystyle(1\!-\!\delta l_1l_2)(1\!-\!\delta^2 l_1 l_2)} \times p.
  \end{array}
\end{equation}

@@

In the case of limited payoffs memory both the role model and the learner
estimate their fitness after interacting with a single member of the population.
At each time step there are five possible pairings. They interact with each
other with a probability \(\frac{1}{N - 1}\), and thus they do not interact with
other with a probability \(1 - \frac{1}{N - 1}\). In the latter case, each of
them can interact with either a mutant or a resident. Both of them interact with
a mutant with a probability $\frac{(k-1)(k-2)}{(N-2)(N-3)}$ and both interact
with a resident with a probability $\frac{(N-k-1)(N-k-2)}{(N-2)(N-3)}$. The last
two possible pairings are that either of them interacts with a resident whilst
the other interacts with a mutant, and this happens with a probability
$\frac{(N-k-1)(k-1)}{(N-2)(N-3)}$. Thus we the probability that the randomly
chosen resident obtained a payoff of $u_R$ in the last round of his respective
game, and that the mutant obtained a payoff of $u_M$ as $x(u_R,u_M)$.

\begin{equation}\label{eq:Chi}
\setlength{\arraycolsep}{1pt}
\begin{array}{llrl}
x(u_1,u_2)	 &=&\displaystyle \frac{1}{N\!-\!1}\cdot  &v_{u_1}(s_1,s_2)\cdot 1_{(u_1,u_2)\in \mathcal{U}^2_F}\\[0.5cm]
&+	
&\displaystyle \left(1\!-\!\frac{1}{N\!-\!1}\right)  
&\left[ \frac{k\!-\!1}{N\!-\!2}\frac{k\!-\!2}{N\!-\!3} v_{u_1}(s_1,s_2) v_{u_2}(s_2,s_2) + 
 \frac{k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!1}{N\!-\!3} v_{u_1}(s_1,s_2) v_{u_2}(s_2,s_1)\right.\\[0.5cm]
&&&\left. +\frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{k\!-\!1}{N\!-\!3} v_{u_1}(s_1,s_1) v_{u_2}(s_2,s_2) + 
 \frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!2}{N\!-\!3} v_{u_1}(s_1,s_1) v_{u_2}(s_2,s_1)\right].
\end{array}
\end{equation}

The first term on the right side corresponds to the case that the learner and
the role model happened to be matched during the game stage, which happens with
probability $\frac{1}{(N\!-\!1)}$. The probability that the number of mutants
increases and decreases by one in the case of limited memory are now given by,

\begin{align}\label{eq:limited_memory_lambdas}
\lambda^+_k=\frac{N\!-\!k}{N} \frac{k}{N} \sum_{u_{R},u_{M}\in\mathcal{U}} x(u_{R},u_{M}) \rho(u_{R},u_{M}) \quad \text{and} \quad
\lambda^-_k=\frac{N\!-\!k}{N} \frac{k}{N} \sum_{u_{R},u_{M}\in\mathcal{U}} x(u_{R},u_{M}) \rho(u_{M},u_{R}).
\end{align}

@@

\subsubsection*{Simulation Results}

>>>>

A pairwise comparison process starts with assigning all
individuals of the population the same strategy. Thenceforth each elementary time step of
the process consists of the mutation phase, the game phase and the update phase.
In the game phase individuals are matched in pairs and that they participate in
a repeated 2 person donation game; a special case of the prisoner's dilemma. In
the donation game there are two actions: cooperation (\(C\)) and defection
(\(D\)). By cooperating a player provides a benefit \(b\) to the other player at
their cost \(c\), with \(0 < c < b\). Thus the payoffs for a player in each turn
are given by,

@@

Let \(\mathbf{u} = (b-c, -c, b, 0)\) be the round payoffs in a vector format, and let
\(\mathcal{U} = \{r, s, t, p\}\) denote the set of feasible payoffs, where \(r\)
denotes the payoff of mutual cooperation, \(s\) the sucker's payoff, \(t\) the
temptation to defect payoff, and \(p\) the punishment payoff.

In repeated games there are infinite many strategies, however, similar to the
literature we will assume that individuals use reactive strategies. A reactive
strategy considers only the previous action of the other player, and thus, a
reactive strategy \(s\) can be written as a three-dimensional vector \(s=(y, p,
q)\). The parameter \(y\) is the probability that the strategy opens with a
cooperation and \(p\), \(q\) are the probabilities that the strategy cooperates
given that the opponent cooperated and defected equivalently. The play between a
pair of reactive strategies $s_1 = (y_1, p_1, q_1)$ and $s_2 = (y_2, p_2, q_2)$
can be model as a Markov process with the transition matrix \(M\),

@@

For the results presented here we assume that mutations are rare (\(\mu
\rightarrow 0\)). In fact, so rare that only two different strategies can be
present in the population at any given time. However, in the Supplementary Information
Section 9 we show that the main result holds for \(\mu \neq 0\).
The case of low mutation is vastly adopted because it allows us to explicitly
calculate the fixation probability of a newly introduced mutant. More specifically,
at each step
one individual adopts a mutant strategy randomly selected from the set of
feasible strategies. The fixation probability \(\phi_{M}\) of the mutant
strategy can be calculated explicitly as,

@@

where \(k\) is the number of
mutants and \(\lambda^-_k, \lambda^+_k\) are the probabilities that the number of
mutants decreases and increases respectively.
Depending on the fixation probability \(\phi_{M}\) the mutant either
fixes (becomes the new resident) or goes extinct. Regardless, in the next elementary
time step another mutant strategy is introduced to the population. We iterate
this elementary population updating process for a large number of mutant
strategies and we record the resident strategies at each time step. The
probabilities \(\lambda^-_k \text{ and } \lambda^+_k\) depend on the fitness
of the mutant and the resident strategies. In the next section we
present how fitness is calculated in the cases of perfect and limited payoff memory.

\subsection{Fitness based on Perfect and Limited Payoff Memory}

In the perfect payoff memory case an individual updates based on the average payoff
against each other member of the population, otherwise known as expected payoffs. 
The payoff of a reactive strategy \(s_1\) against the reactive strategy \(s_2\)
in an infinitely repeated game ($\delta \rightarrow 1$) is explicitly
calculated as,

\[\langle\mathbf{v}(s_1,s_2),\mathbf{u}\rangle.\]

In a population of size \(N\) there are \(k\) mutants and \(N - k\) residents.
Let \(s_M =(y_M, p_M, q_M)\) and \(s_R = (y_R, p_R, q_R)\) denote the
strategies of a mutant and a resident, the expected payoffs \(\pi_R\) and
\(\pi_M\) are given by,

@@

The probabilities that the number of mutants decreases and increases,
\(\lambda^-_k\) and \(\lambda^+_k\), in the perfect payoff memory case are
defined as,

@@

In the case of limited payoff memory we initially define the probability that a
reactive strategy receives the payoff $u\!\in\! \mathcal{U}$ in the very last
round of the game. This is given by Proposition~\ref{proposition:last_round}
(see Supplementary Information Section 2.2.1 for proof).

@@

  v_{r}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{y_1y_2}{1\!-\!\delta^2 l_1 l_2}+\delta \frac{\Big(q_1+l_1\big((1\!-\!\delta)y_2+\delta q_2\big)\Big) \Big(q_2+l_2\big((1\!-\!\delta)y_1+\delta q_1\big)\Big)}
  {\displaystyle(1\!-\!\delta l_1l_2)(1\!-\!\delta^2 l_1 l_2)},\\[1cm]

  v_{s}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{y_1\bar{y}_2}{1\!-\!\delta^2 l_1 l_2}+\delta \frac{\Big(q_1+l_1\big((1\!-\!\delta)y_2+\delta q_2\big)\Big) \Big(\bar{q}_2+\bar{r}_2\big((1\!-\!\delta)y_1+\delta p_1\big)\Big)}
  {\displaystyle(1\!-\!\delta l_1l_2)(1\!-\!\delta^2 l_1 l_2)},\\[1cm]

  v_{t}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{\bar{y}_1y_2}{1\!-\!\delta^2 l_1 l_2}+\delta \frac{\Big(\bar{q}_1+\bar{r}_1\big((1\!-\!\delta)y_2+\delta p_2\big)\Big) \Big(q_2+l_2\big((1\!-\!\delta)y_1+\delta q_1\big)\Big)}
  {\displaystyle(1\!-\!\delta l_1l_2)(1\!-\!\delta^2 l_1 l_2)},\\[1cm]

  v_{p}(s_1,s_2) &= &\displaystyle (1\!-\!\delta)\frac{\bar{y}_1\bar{y}_2}{1\!-\!\delta^2 l_1 l_2}+\delta \frac{\Big(\bar{q}_1+\bar{r}_1\big((1\!-\!\delta)y_2+\delta p_2\big)\Big) \Big(\bar{q}_2+\bar{r}_2\big((1\!-\!\delta)y_1+\delta p_1\big)\Big)}
  {\displaystyle(1\!-\!\delta l_1l_2)(1\!-\!\delta^2 l_1 l_2)}.
  \end{array}
\end{equation}

@@

In the case of limited payoffs memory both the role model and the learner
estimate their fitness after interacting with a single member of the population.
At each time step there are five possible pairings. They interact with each
other with a probability \(\frac{1}{N - 1}\), and they do not interact with
other with a probability \(1 - \frac{1}{N - 1}\). In the latter case, each of
them can interact with either a mutant or a resident. Both of them interact with
a mutant with a probability $\frac{(k-1)(k-2)}{(N-2)(N-3)}$ and both interact
with a resident with a probability $\frac{(N-k-1)(N-k-2)}{(N-2)(N-3)}$. The last
two possible pairings are that either of them interacts with a resident whilst
the other interacts with a mutant, and this happens with a probability
$\frac{(N-k-1)(k-1)}{(N-2)(N-3)}$. We define the probability that the randomly
chosen resident obtained a payoff of $u_R$ in the last round of his respective
game, and that the mutant obtained a payoff of $u_M$ as $x(u_R,u_M)$.

\begin{equation}\label{eq:Chi}
\setlength{\arraycolsep}{1pt}
\begin{array}{llrl}
x(u_R,u_M)	 &=&\displaystyle \frac{1}{N\!-\!1}\cdot  &v_{u_R}(s_R,s_M)\cdot 1_{(u_R,u_M)\in \mathcal{U}^2_F}\\[0.5cm]
&+	
&\displaystyle \left(1\!-\!\frac{1}{N\!-\!1}\right)  
&\left[ \frac{k\!-\!1}{N\!-\!2}\frac{k\!-\!2}{N\!-\!3} v_{u_R}(s_R,s_M) v_{u_M}(s_M,s_M) + 
 \frac{k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!1}{N\!-\!3} v_{u_R}(s_R,s_M) v_{u_M}(s_M,s_R)\right.\\[0.5cm]
&&&\left. +\frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{k\!-\!1}{N\!-\!3} v_{u_R}(s_R,s_R) v_{u_M}(s_M,s_M) + 
 \frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!2}{N\!-\!3} v_{u_R}(s_R,s_R) v_{u_M}(s_M,s_R)\right].
\end{array}
\end{equation}

The first term on the right side corresponds to the case that the learner and
the role model happened to be matched during the game stage, which happens with
probability $\frac{1}{(N\!-\!1)}$.

The probability that the number of mutants increases and decreases by one in the
case of limited payoff memory are now given by,

\begin{align}\label{eq:limited_memory_lambdas}
\lambda^+_k=\frac{N\!-\!k}{N} \cdot \frac{k}{N} \cdot \sum_{u_{R},u_{M}\in\mathcal{U}} x(u_{R},u_{M}) \rho(u_{R},u_{M}) \quad \text{and} \quad
\lambda^-_k=\frac{N\!-\!k}{N} \cdot \frac{k}{N} \cdot \sum_{u_{R},u_{M}\in\mathcal{U}} x(u_{R},u_{M}) \rho(u_{M},u_{R}).
\end{align}

@@

\section{Simulation Results}
